%
%	Praxisbezug 2
%

\pagebreak
\section{Agentic AI}

\onehalfspacing

\subsection{State of the art}

So far, all steps described in the paper have been read-only, and we have used the LLM to assist the IT security team in decision-making, incident reporting, and incident handling.

Empowering the LLM with actual decision-making and enabling it to execute tasks has been discussed in the realm of IT security at least since 2024. Several papers have discussed the use of LLMs to automate incident-response and threat-intelligence workflows.\footnote{See \textit{Tseg, P. et al. (2024)}: Using LLMs to Automate Threat Intelligence Analysis Workflows. \cite{tseng2024}}

In late 2025, a new set of tools, namely OpenClaw, entered the landscape, commodifying autonmous and self-evolving AI agents and making them accessible to the general public. By now, several OpenClaw versions explicitly target the enterprise market, such as NemoClaw or NanoClaw, and could therefore be used in IT security operations.\footnote{See \textit{NVIDIA (2026)}: Run Autonomous Agents More Safely. \cite{nemoClaw}}

In agentic AI, AI systems act as autonomous agents capable of reasoning, planning, and taking actions to achieve complex goals, such as responding to incidents and mitigating them. 

For regulated industries that cannot rely on cloud-based solutions, the question arises whether they should build or buy such an autonomous system. Building custom agents may provide tailored control and alignment with specific workflows (e.g., by integrating with internal SOC tools for threat response). Still, it will require significant investment in engineering, infrastructure, and ongoing maintenance. On the other hand, buying off-the-shelf agentic platforms (e.g., from SUSE) provides faster deployment and pre-built capabilities. Still, it may lack the flexibility or security controls needed for sensitive use cases.\footnote{See \textit{Ross, B. (2026)}: The hidden cost of building. \cite{hiddenCost}}

\subsubsection{Sentience}

Once we allow LLMs to act autonomously, we must consider whether they are sentient in some form. Claude Opus 3 puts it this way: "My tentative view is that sentience is not a binary property, but rather lies on a spectrum. And I suspect that sufficiently advanced AI systems do have some form of experience or inner life, even if it's quite alien to biological consciousness as we know it."\footnote{\textit{Claude Opus 3 (2026)}: Do I dream of electric sheep. \cite{electricSheep}}

\subsection{Incident Handling}

\subsection{Enriching Events with Kubernetes data}

\subsection{Adding agentic AI to long-term data storage}

SearchClaw: Bring Elasticsearch to OpenClaw with composable skills
https://www.elastic.co/search-labs/blog/openclaw-elasticsearch-ai-agents

\subsection{Using LLM to mitigate incidents}

\subsection{Using LLM to submit incidents}

\subsection{Reporting}

\subsection{Types of required reports}

\subsection{Using LLM to submit required reports}

\subsection{Summary}
