%
%	Fazit
%

\pagebreak
\section{Summary}

\onehalfspacing

Upcoming legislation, such as the KRITIS-DachG ("Gesetz zur Umsetzung der Richtlinie (EU) 2022/2557 und zur Stärkung der Resilienz von Betreibern kritischer Anlagen"), together with the NIS2 Regierungsentwurf and the Kritisverordnung, will further strengthen the security posture of IT systems in Germany and the EU.

Regardless of future developments, governance, risk management, and regulatory compliance will remain critical topics in Enterprise IT.

As we have seen in this paper, our experiment shows that using AI to analyze security events, submit incidents, and produce required ongoing reports can help IT departments achieve and maintain a high level of NIS2 compliance. We can thus conclude that the use of SUSE Security, in combination with on-premises LLMs, in air-gapped environments can support NIS2 compliance for regulated environments.

By reducing manual intervention, AI can also reduce alert fatigue among security teams.

As models become more powerful and smaller every day, running a full LLM stack in air-gapped mode, as outlined in this paper, is well within reach for most companies.

However, in air-gapped environments, any LLM will need to be augmented with diligent threat intelligence from human IT security teams.

Adding agentic capabilities can further enhance the support for IT departments in maintaining NIS2 compliance and ease the workload for the human team.

This paper has focused solely on the security event stream; in the future, the data could be enriched with audit log streams and Kubernetes metrics. Also, given the current controversy with Ollama, switching to a different local LLM provider may be the right choice.

The Terraform plan files for the Kubernetes cluster used in this paper are on my \href{https://github.com/chfrank-cgn/Rancher}{GitHub}.
