%
%	Begrifflichkeiten
%

\pagebreak
\section{NIS2 Incident Handling \& Reporting}

\onehalfspacing

\subsection{The Network and Information Security Directive}

The European Commission and the High Representative of the Union for Foreign Affairs and Security Policy presented a new EU Cybersecurity Strategy at the end of 2020, the second version of the Network and Information Security Directive (NIS2).\footnote{See \textit{EU (2023)}: Cybersecurity Policies. \cite{cyberPol}} NIS2 is a new EU cybersecurity law that will make a significant difference. The goal of NIS2 is to improve the overall level of cybersecurity across the European Union.

NIS2 will:

\begin{itemize}
    \item Apply to a wide range of organizations, from big tech companies to small businesses
    \item Focus on critical sectors like energy, transportation, healthcare, and finance
    \item Introduce rules covering things like risk management, incident reporting, and supply chain security
    \item Carry hefty fines for non-compliance, so companies will need to take it seriously
\end{itemize}

We will first examine the key components and organizations newly introduced in relation to the directive, and then focus on the incident-handling and reporting requirements.

\subsubsection{EU Cybersecurity Agency ENISA}

The European Union Agency for Cybersecurity (ENISA) is a key player in ensuring high Cybersecurity across Europe. It is the EU's main cybersecurity organization, helping keep all member countries safe from cyber threats.

Established in 2004 and empowered by the EU Cybersecurity Act, ENISA contributes to EU cyber policy, promotes trustworthy Information and Communications Technology products and services, collaborates with Member States and EU bodies, and prepares Europe for future cyber threats. 

Here's what ENISA does:

\begin{itemize}
    \item They help EU countries work together on cybersecurity issues
    \item Create guidelines and best practices for staying safe online
    \item Research new cyber threats and share information about them
    \item Help train cybersecurity professionals across Europe
    \item Support the implementation of laws like NIS2
\end{itemize}

By sharing knowledge, building capacity, and raising awareness, ENISA works with stakeholders to strengthen trust in the connected economy, boost resilience, and safeguard Europe's society and citizens in the digital age.\footnote{See \textit{EU (2025)}: Who we are. \cite{aboutEnisa}}

ENISA plays a huge role in helping countries understand and implement NIS2 requirements. They produce guidance documents, organize workshops, and serve as the EU's go-to resource for cybersecurity knowledge.\footnote{See \textit{ENISA (2025)}: Technical Implementation Guidance. \cite{enisaTech}}

They also help coordinate responses to large-scale cyber incidents across multiple countries.

\subsubsection{EU Cybersecurity Act}

The EU Cybersecurity Act was adopted in 2019 and is the central piece of legislation strengthening the EU's cybersecurity capabilities. It enhances the role of ENISA, the EU Agency for Cybersecurity, and establishes a robust certification framework for products and services.

Key provisions of the EU Cybersecurity Act:

\begin{itemize}
    \item Strengthened ENISA: The Act grants ENISA a permanent mandate, increases its resources, and assigns it new responsibilities.
    \item Certification framework: ENISA will play a pivotal role in establishing and maintaining a European cybersecurity certification framework.
    \item Public information: ENISA will provide information on certification schemes and issued certificates through a dedicated website.
    \item Operational cooperation: ENISA will facilitate cooperation between EU Member States in handling cybersecurity incidents and coordinating responses to large-scale cross-border attacks.\footnote{See \textit{EU (2023)}: The EU Cyber Security Act. \cite{cyberSec}}
\end{itemize}

The EU Cybersecurity Act works well together with NIS2. While NIS2 focuses on protecting critical infrastructure, the Cybersecurity Act makes sure the products and services used are actually secure in the first place.

\subsubsection{EU Cyber Resilience Act}

The Cyber Resilience Act (CRA) is a significant step towards safeguarding consumers and businesses from cybersecurity vulnerabilities. It is about ensuring that products with digital elements are secure from the outset. 

By introducing mandatory cybersecurity requirements for manufacturers and retailers, the CRA addresses inadequate security features in products and software. It also aims to empower consumers and businesses by providing clear information on product cybersecurity and ensuring that products are designed and maintained with security in mind. 

The CRA's key benefits include:

\begin{itemize}
    \item Harmonized rules: Consistent rules for products with a digital component.
    \item Cybersecurity framework: Requirements for planning, design, development, and maintenance of products.
    \item Duty of care: Obligation to ensure product security throughout its lifecycle.
    \item Consumer protection: Improved ability for consumers to make informed choices.
    \item Business security: Enhanced protection for businesses using digital products.\footnote{See \textit{EU (2024)}: Cyber Resiliency Act. \cite{cyberResiliency}}
\end{itemize}

The CRA also mandates Security by Design by requiring companies to:

\begin{itemize}
    \item Do risk assessments
    \item Fix vulnerabilities throughout the product's entire lifecycle
    \item Report serious incidents to authorities
    \item Provide security updates for a reasonable time period
\end{itemize}

The CRA was adopted in 2024, and companies will have approximately 3 years to prepare. 

As with NIS2, companies that fail to comply may incur penalties of up to 2.5\% of their global revenue.

\subsubsection{EU Cyber Solidarity Act}

The goal of this act is to foster solidarity: When a major cyberattack hits one EU country, other countries can respond to help.

Why is solidarity so important regarding cybersecurity?

\begin{itemize}
    \item Cyber attacks don't care about borders; they can spread from one country to another super quickly
    \item Some attacks are so big that even large countries need help dealing with them
    \item By working together, the countries in the EU become much stronger against cyber criminals and other hostile actors
\end{itemize}

To achieve this, the act seeks to improve preparedness, detection, and response to significant cybersecurity threats by establishing a European Cybersecurity Alert System and a comprehensive Cybersecurity Emergency Mechanism.

The key features of the EU Cyber Solidarity Act are:

\begin{itemize}
    \item European Cybersecurity Alert System: This system will consist of interconnected Security Operation Centers (SOCs) across the EU, using advanced technologies like AI and data analytics to detect and share threat warnings.
    \item Cybersecurity Emergency Mechanism: This mechanism will provide a framework for coordinated responses to large-scale cyberattacks and crises.
    \item Enhanced capacities: The Act will strengthen the EU's capacity to detect, prepare for, and respond to cybersecurity incidents.
\end{itemize}

The Cyber Solidarity Act is still relatively new; it was proposed in 2023 and is now progressing through the EU legislative process.

The European Cybersecurity Shield is a component of the act. It is a pilot initiative involving three cross-border Security Operations Center consortia. Launched in November 2022, this pilot project aims to test and refine the concept of a networked system of SOCs for early threat detection and sharing.\footnote{See \textit{EU (2024)}: Cyber Solidarity Act. \cite{cyberSol}}

\subsubsection{EU Cybersecurity}

We've covered the four acts that build the complete EU cybersecurity framework:

\begin{itemize}
    \item NIS2: Protects critical infrastructure
    \item Cybersecurity Act: Certifies products and strengthens ENISA
    \item Cyber Resilience Act: Makes products secure by design
    \item Cyber Solidarity Act: Countries help each other during attacks
\end{itemize}

We now examine the national counterpart of the EU institutions.

\subsection{Bundesamt für Sicherheit in der Informationstechnik}

The BSI in Germany refers to the Bundesamt für Sicherheit in der Informationstechnik (Federal Office for Information Security). It is Germany's main cybersecurity command center, analogous to ENISA at the EU level, and was established in 1991.

The BSI is headquartered in Bonn and operates under the Federal Ministry of the Interior. It plays a crucial role in Germany's national cybersecurity strategy and works closely with international partners such as NIST on cybersecurity matters.

As part of their role, the BSI:

\begin{itemize}
    \item Protects Germany's critical infrastructure, such as power plants and hospitals
    \item Handles the KRITIS (critical infrastructure) regulations
    \item Issues cybersecurity guidelines and standards for government and businesses
    \item Publishes the famous "IT-Grundschutz" (IT baseline protection) standards that a lot of companies use
    \item Investigates cyber incidents and coordinates responses
    \item Certifies IT products for security
    \item Trains cybersecurity professionals
    \item Operates its own cybersecurity research lab
\end{itemize}

Among other things, a draft bill amending the BSI Act (BSIG) expands the group of regulated organizations to include important and particularly important institutions.\footnote{See \textit{BSI (2025)}: NIS-2-Regierungsentwurf. \cite{presseNis2}}

In its new role, the BSI is basically Germany's national authority for implementing and enforcing NIS2 and will, in the future, supervise approximately 29,500 institutions subject to new legal obligations in the area of IT security.

\subsection{NIS2 Articles 21 and 23}

Returning to the NIS2 directive, much of its content concerns reporting requirements and EU-wide cooperation and institutions, which we will partly analyze further in this paper.\footnote{See \textit{EU (2022)}: NIS2 Directive. \cite{nis2}}

The NIS2 directive itself is a legal document organized into Chapters and Articles. It has a pan-European scope and targets critical businesses. NIS2 applies to many entities that provide essential services to the European economy and society, such as Operators of Essential Services (OES) and Digital Service Providers (DSPs). While NIS2 primarily targets medium and large enterprises, smaller entities may still be affected, depending on the services they provide.

Article 21 defines the required cybersecurity risk-management measures for critical infrastructure. In paragraph 2, point (g), NIS2 calls for basic cyber hygiene practices and cybersecurity training.

Any entity covered by the directive must, thus, have an IT Security Policy to implement basic cyber hygiene.\footnote{See \textit{NIS 2 Compliant.org (2024)}: List of Policies Required by NIS 2 Directive. \cite{nisPols}} To strengthen their security postures, entities could rely on one of the major frameworks and standards, such as the NIST SP 800 series, ISO/IEC 27001, Mitre Att\&ck, or CIS Controls.\footnote{See \textit{NIS 2 Compliant.org (2024)}: Requirements Checklist for NIS 2. \cite{nisReqs}}

NIS2 is a legal framework, and compliance will be mandatory for European critical entities, ensuring widespread adoption. Unlike the CISecurity or ISO 27001 frameworks, adherence to the NIS2 Directive's requirements is not optional for entities classified as important or particularly important; it will be mandated by law.

In addition to Article 21, the NIS2 Directive also contains Article 23, which outlines the requirements for EU-wide incident reporting. Article 23 defines what constitutes an incident, the mandatory reports, and the content required in these reports. With pan-European reporting of cyber attacks, a coordinated response should become much more manageable.

\subsection{Threat Intelligence}\label{threat-intelligence}

Threat Intelligence comprises the collection, analysis, and sharing of information on current and potential cyber threats. NIS2 doesn't explicitly require entities to have threat intelligence, but it does require companies to implement robust cybersecurity risk management measures.

Threat intelligence comprises three stages that recur daily.

Data Collection: Gathering info from sources like
\begin{itemize}
    \item Security research reports
    \item Government advisories
    \item Previous attack patterns
    \item Honeypots
\end{itemize}

Analysis and Context: Taking all that raw data and making sense of it:
\begin{itemize}
    \item Who are the attackers?
    \item What are their motivations?
    \item What techniques are they using?
    \item Which industries are they targeting?
\end{itemize}

Actionable Insights: Turning that analysis into actionable insights:
\begin{itemize}
    \item There's a new ransomware targeting healthcare systems
    \item Patch this vulnerability, it's being actively exploited
    \item Watch out for phishing emails with this specific signature
\end{itemize}

Threat intelligence enables organizations to prepare for and protect against threats.

\subsection{Incident Handling requirements}

NIS2 includes comprehensive requirements for incident handling to ensure organizations are prepared to detect, respond to, and recover from cybersecurity incidents. These requirements are primarily set out in Articles 21 (Risk Management Measures) and 23 (Incident Reporting), but are supported by broader obligations throughout the directive.

NIS2 emphasizes a proactive and structured approach to incident handling, requiring organizations to implement policies, procedures, and technical measures to manage incidents effectively.

Under Article 21 of NIS2, entities (both essential and important) must establish incident-handling processes as part of their risk management measures. Specifically, organizations must:

\begin{itemize}
    \item Develop and maintain an incident response plan
    \item Clearly identify teams (e.g., incident response team, IT, legal, PR) and their duties
    \item Have formalized escalation procedures and communication protocols
    \item Perform regular incident response drills
\end{itemize}

Organizations need threat intelligence integrated into their teams to remain aware of emerging risks.

Furthermore, organizations must implement these technical and organizational measures to detect incidents promptly:

\begin{itemize}
    \item Intrusion Detection Systems (IDS)
    \item Security Information and Event Management (SIEM)
    \item Log monitoring and anomaly detection
\end{itemize}

Once a security incident occurs, it must be classified by severity and impact (e.g., low, medium, high, critical).

Organizations need to define strategies to contain incidents and prevent further damage, and later to remove the root cause of the incident (e.g., by removing malware or closing vulnerabilities).

After the incident, the teams involved should conduct a post-incident review and identify lessons learned and improvements for future incident handling.

\subsection{Reporting requirements}

NIS2 Article 23 addresses the obligations to report incidents for entities covered by the NIS2 Directive.\footnote{See \textit{EU (2022)}: NIS2 Directive. \cite{nis2}}

This is a critical part of the directive, as it mandates the manner and timing by which organizations must report cybersecurity incidents to authorities. Article 23 outlines the detailed obligations for reporting incidents to competent authorities (e.g., the BSI). 

The goal is to ensure timely awareness of cybersecurity threats and incidents across all EU member countries and to enable rapid response and mitigation.

All entities covered by NIS2 must also report cybersecurity incidents.

They must report incidents that have a significant impact on the security of their network and information systems. An incident is reportable if it disrupts or threatens to disrupt the provision of services or if it impacts the security of the network or information systems (e.g., data breaches, ransomware attacks, DDoS attacks). Reports also need to include information on cross-border implications (Article 24).

There is a strict timeline in which these reports should occur:

\begin{itemize}
    \item Within 24 hours, a preliminary notification that an incident has occurred needs to be sent
    \item Within 72 hours, a detailed report with a description of the incident and its impact on services
    \item Within a month, a detailed report with root cause analysis and lessons learned
\end{itemize}

Failure to comply with Article 23 can result in administrative fines of up to 2\% of global annual turnover for essential entities and up to 1.4\% of global annual turnover for important entities.

In the next chapter, we will describe our experiment and the setup for processing Kubernetes security events, and assess whether a local LLM could support IT security teams in fulfilling their obligations under the NIS2 directive to handle incidents and report them.
